Detecting skill level and mental template in biface morphology: Archaeological and experimental insights

### 2022-06-06 manuscript/manuscript.Rmd
added the paragraph of mental template.

---
title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, encompassing raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

@@

Mental template: short literature review

Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data? 2) What is the effect of training on these two aspects?

@@

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples prepared by R. I. [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006]. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].

>>>>

---
title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic still remain elusive. Among many factors intentified to contribute to the considerable variation of biface morphology, skill level and mental template have been frequently quoted and discussed. In this paper, we \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept essentially lies in the very foundation of the cultural historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial temporal framework. Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of biface at regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental template (or "imposed form") has been recongnized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraint [@white1995], and the lacking of rigorous studies of its corresponding cognitive processes. Until very recently, several researchers have actively addressed these critiques and reconceuptualized the mental template in the study of biface morphology. Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Further, raw material is often treated as a crucial variable to be tested in the very beginning of a research design focusing on mental template [@garcía-medrano2019]. In terms of the cognitive mechanisms behind mental template, Ho and collegues [-@ho2022] recently developed a series of nevigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors. In short, when exercized with proper caution, mental template still has its value in our study of biface morphological variation.

@@

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers. Since the novice biface collection is generated from an 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data? 2) How does training affect novices' performance on these two aspects?

@@

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples prepared by R. I. [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006], which also at least alleviate the effect of resharpening process to some extent. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].
